\Ace: Growing a Statically-Typed Language Inside a Python

### 2013-11-14 ace-pldi14/ace-pldi14.tex
pldi14 2.5 p2+3

To produce an output file from an Ace compilation script like \verb|listing|\texttt{\ref{compscript}}\verb|.py|, the command \verb|acec| can be invoked from the shell, as shown in Listing \ref{mapc}. The \verb|acec| compiler operates in two steps:
\begin{enumerate}
\item Executes the provided Python file (\verb|listing3.py|).
\item Extracts the translation from concrete functions and other top-level constructs (e.g. types requiring declarations, or generated imports and pragmas) in the top-level Python environment.  This may produce one or more files (here, just \verb|listing3.cl|) as determined by the active targets that were used (for example, a web framework built upon Ace might produce separate HTML, CSS and JavaScript files; see Sec. \ref{active-targets}).
\end{enumerate}

@@

\subsection{Types}
\begin{codelisting}
\lstinputlisting{listing4.py}
\caption{[\texttt{listing\ref{compscript}.py}] The generic \texttt{map} function compiled to map the \texttt{negate} function over two  types of input.}
\label{compscript}
\end{codelisting}
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec listing3.py`
Hello, compile-time world!
[ace] TypeError in listing1.py (line 6, col 28): 
      'GenericFnType(negate)' does not support [].
[acec] listing3.cl successfully generated.
\end{lstlisting}
\caption{Compiling \texttt{listing\ref{compscript}.py} using the \texttt{acec} compiler.}
\label{mapc}
\end{codelisting}
\begin{codelisting}
\lstinputlisting[style=OpenCL]{listing5.cl}
\caption{[\texttt{listing\ref{compscript}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
\label{mapout}
\end{codelisting}
Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned on lines \ref{compscript}.7-\ref{compscript}.10. In Ace, types are themselves values that can be manipulated at compile-time, in contrast to most  contemporary languages, where user-defined types are declaratively written at compile-time but cannot be constructed, inspected or passed around programmatically. More specifically, types are instances of a Python class that implements the \verb|ace.ActiveType| interface (see Sec. \ref{extensions}). Python, then, serves as the \emph{type-level language} of Ace. 
Being Python values, types can be assigned to variables for convenience (a construct similar to \verb|typedef| in C or \verb|type| in Haskell). Types, like all compile-time mechanisms in Ace (functions, bases, etc.), do not have mutable state and operate in a referentially transparent manner.% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.

The type named \verb|T1| on line \ref{compscript}.7 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four memory spaces defined by OpenCL \cite{opencl11}). It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to an address space (here, \verb|clx.global_|) and the type being pointed to, \verb|clx.float|, the Ace type corresponding to \verb|float| in OpenCL. 
The \verb|clx| library contains a full implementation of the OpenCL type system. We will return to how types are implemented in Section \ref{att}. We note that we are being more verbose than necessary for the sake of pedagogy. The \verb|clx| library includes a concise shorthand for OpenCL's built-in types: \verb|T1| could be written \verb|clx.gp(clx.f32)|. We also note that Ace is \emph{unopinionated} about issues like memory safety. We will show how to implement safer alternatives to raw pointers in Sec. \ref{examples}, but we allow users to choose a low level of abstraction if they wish (e.g. for compatibility with existing code).

The type \verb|T2| on line 8 does not correspond to an OpenCL type directly. Rather, it is a global pointer to a user-defined primitive type representing complex numbers (integers, in this case). OpenCL does not include primitive support for complex numbers (as of this writing)

@@

Running \verb|acec listing3.py| would produce no output. To create a {\em concrete function} (that is, an instance of the class \verb|ace.ConcreteFn|) that can be emitted by the compiler, types must be assigned to each of the arguments. Listing \ref{compscript} shows how to use the \verb|compile| method to specialize \verb|map| in \emph{two} different ways to apply the \verb|add5| function, defined on Lines 4-6, to arrays  that reside in global memory (OpenCL associates a memory space with pointer types). Line 9 results in a version specialized for arrays of \verb|double|s and Line 10 results in a version for arrays of \verb|int|s. The output of compilation is shown in Listing \ref{mapout}.

The \verb|map| function defined in Listing \ref{map} will compile to one or more kernels that applies a transfer function, \verb|f|, to the element of the input array, \verb|input|, corresponding to its global id. It writes the result of this call into the corresponding location in the provided output array, \verb|output|. If the kernel is running in thread 0, it prints a message. The functions \verb|get_global_id| and \verb|printf| are OpenCL primitive functions. To emphasize again that the semantics of this code differs from Python's, we note that these functions have no meaning in Python.

%\begin{codelisting}
%\lstinputlisting{hello.py}
%\caption{\texttt{[hello.py]} A basic Ace program demonstrating the two-phase structure of Ace programs and libraries.}
%\label{hello}
%\end{codelisting}
%A variant of the standard ``Hello, World!'' example written in Ace is shown in Listing \ref{hello} and its compilation to statically-typed OpenCL kernel code is demonstrated in Listing \ref{helloout}. The OpenCL kernel language is a variant of C99 with some additions and restrictions to facilitate execution on GPUs and other accelerators, in addition to conventional CPUs. We will discuss it further in subsequent examples.
%We emphasize that this module, which we use throughout the paper, is simply a library like any other. The core of Ace gives no special treatment to it; it is distributed together with Ace for convenience. Aspects of its implementation will be described  in Section \ref{att}.

This example demonstrates several key design decisions that characterize Ace: static typing of run-time behavior, a Python-based syntax, a phase distinction between compile-time and run-time logic, and programmatic

@@

\section{Extensibility}\label{att}
If no active target is explicitly provided, as in our example, the active base  provides a default target. The default target provided by \verb|clx.base| is \verb|clx.opencl|, but other targets, like \verb|clx.cuda|, a vendor-specific language similar to OpenCL, as well as Python itself (\verb|ace.python|), can also be defined in libraries. One base can support multiple targets. We will discuss targets in Sec. \ref{targets}.

>>>>

To produce an output file from an Ace compilation script like \verb|listing|\texttt{\ref{compscript}}\verb|.py|, the command \verb|acec| can be invoked from the shell, as shown in Listing \ref{mapc}. The \verb|acec| compiler operates in two steps:
\begin{enumerate}
\item Executes the provided Python file (\verb|listing3.py|).
\item Extracts the translation from concrete functions and other top-level constructs (e.g. types requiring declarations, or generated imports and pragmas) in the top-level Python environment.  This may produce one or more files (here, just \verb|listing3.cl|) as determined by the active targets that were used (for example, a web framework built upon Ace might produce separate HTML, CSS and JavaScript files; see Sec. \ref{atargets}).
\end{enumerate}

@@

\subsection{Types}
\begin{codelisting}
\lstinputlisting{listing4.py}
\caption{[\texttt{listing\ref{compscript}.py}] The generic \texttt{map} function compiled to map the \texttt{negate} function over two  types of input.}
\label{compscript}
\end{codelisting}
\begin{codelisting}
\begin{lstlisting}[style=Bash]
$ `acec listing3.py`
Hello, compile-time world!
[ace] TypeError in listing1.py (line 6, col 28): 
      'GenericFnType(negate)' does not support [].
[acec] listing3.cl successfully generated.
\end{lstlisting}
\caption{Compiling \texttt{listing\ref{compscript}.py} using the \texttt{acec} compiler.}
\label{mapc}
\end{codelisting}
\begin{codelisting}
\lstinputlisting[style=OpenCL]{listing5.cl}
\caption{[\texttt{listing\ref{compscript}.cl}] The OpenCL file generated by Listing \ref{mapc}.}
\label{mapout}
\end{codelisting}
Lines \ref{compscript}.3-\ref{compscript}.5 construct the types assigned on lines \ref{compscript}.7-\ref{compscript}.10. In Ace, types are themselves values that can be manipulated at compile-time, in contrast to most  contemporary languages, where user-defined types are declaratively written at compile-time but cannot be constructed, inspected or passed around programmatically. More specifically, types are instances of a Python class that implements the \verb|ace.ActiveType| interface (see Sec. \ref{atypes}). Python, then, serves as the \emph{type-level language} of Ace. 
As Python values, types can be assigned to variables when desired (removing the need for  facilities like \verb|typedef| in C or \verb|type| in Haskell). Types, like all compile-time objects derived from Ace base classes, do not have visible state and operate in a referentially transparent manner (by constructor memoization, which we do not detail here).% These types are all implemented in the \verb|clx| library imported on line 1, none are built into Ace itself.

The type named \verb|T1| on line \ref{compscript}.3 corresponds to the OpenCL type \verb|global float*|: a pointer to a 32-bit floating point number stored in the compute device's global memory (one of four memory spaces defined by OpenCL \cite{opencl11}). It is constructed by applying \verb|clx.Ptr|, which is an Ace type constructor corresponding to pointer types, to a value representing an  address space, \verb|clx.global_|, and the type being pointed to. That type, \verb|clx.float|, is in turn the Ace type corresponding to \verb|float| in OpenCL (which, unlike C99, is always 32 bits). 
The \verb|clx| library contains a full implementation of the OpenCL type system.
We note that we are being more verbose than necessary for the sake of pedagogy. The \verb|clx| library includes a concise shorthand for OpenCL's built-in types: \verb|T1| could be written \verb|clx.gp(clx.f32)|. We also note that Ace is \emph{unopinionated} about issues like memory safety. We will show how to implement safer alternatives to raw pointers in Sec. \ref{examples}\todo{true?}, but we allow users to choose a low level of abstraction if the need arises.

The type \verb|T2| on line \ref{compscript}.4 does not correspond to type in OpenCL directly. It is a pointer to a \emph{complex integer} in global memory. OpenCL does not include primitive support for complex numbers, however -- it is an extension available only in \verb|clx|. The \verb|clx.Cplx| type constructor is parameterized by the numeric type that should be used for the real and imaginary parts, here \verb|clx.int|, the type corresponding to 32-bit OpenCL integers. Arithmetic is supported with other types in the same family, as well as with plain numeric types (treated as if their imaginary part was zero). When targeting OpenCL, Ace expressions of type \verb|clx.Cplx(clx.int)| are compiled to OpenCL expressions of type \verb|int2|, a pair of 32-bit integers. This can be observed in several places on lines \ref{mapout}.14-\ref{mapout}.21. This is merely an implementation detail of \verb|clx.Cplx|, however. An Ace value of type \verb|clx.int2| can\emph{not} be used when a \verb|clx.Cplx(clx.int)| is expected (attempting to do so will result in a static type error). That is, \verb|clx.Cplx| truly extends the type system, it is not merely a type alias (see Sec. \ref{atypes}).

The type \verb|TF| on line \ref{compscript}.5

@@

@@

\section{Extensibility}\label{att}
\subsection{Active Bases}\label{abases}
\subsection{Active Types}\label{atypes}
\subsection{Active Targets}\label{atargets}

If no active target is explicitly provided, as in our example, the active base  provides a default target. The default target provided by \verb|clx.base| is \verb|clx.opencl|, but other targets, like \verb|clx.cuda|, a vendor-specific language similar to OpenCL, as well as Python itself (\verb|ace.python|), can also be defined in libraries. One base can support multiple targets. We will discuss targets in Sec. \ref{targets}.
